Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-03-09 pkg/disttree/inst/draft/draft.Rnw
revised Sections 3.1 and 3.2
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2048 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d2e5153443c5ec375fb5f6b5bed13334c7d77794

\Address{
  Lisa Schlosser, Reto Stauffer, Achim Zeileis \\
  Universit\"at Innsbruck \\
  Department of Statistics \\
  Faculty of Economics and Statistics \\
  Universit\"atsstr.~15 \\
  6020 Innsbruck, Austria \\
  E-mail: \email{Lisa.Schlosser@uibk.ac.at}, \email{Reto.Stauffer@uibk.ac.at},\\
  \email{Achim.Zeileis@R-project.org} \\
  URL: \url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/},\\
  \url{http://retostauffer.org/}, \url{https://eeecon.uibk.ac.at/~zeileis/}\\

@@

In weather forecasting, these properties are appealing in mountainous regions
and complex terrain where small-scale effects are not well-resolved in the NWP
outputs. As these are computed at a coarser scale, there may be abrupt local
effects within a single NWP grid cell. To illustrate this in practice,
precipitation forecasts are obtained with distributional forests at 95
meteorological stations in a mountainous region in the Alps, covering mainly
Tyrol, Austria, and adjacent areas. More specifically, a zero-censored Gaussian
distribution is employed to model 24-hour total precipitation so that the
zero-censored point mass describes the probability of no precipitation on a
given day (see Figure~\ref{axams_onestation}). Forecasts for July are
established based on data from the same month in the years 1985--2012 along with
82~covariates derived from a wide range of NWP ensemble outputs. As
Figure~\ref{axams_onestation} shows, the station-wise models yield a full
distributional forecast for each day -- here for the same day (July~24) at one
station (Axams) over four years (2009--2012) -- based on the previous 24~years
as training data. The corresponding observations conform reasonably well with
the predictions. In Section~\ref{ProbabilisticForecasting} we investigate the
performance of distributional forests in this forecasting task in more detail.
It is shown that they perform at least on par and sometimes clearly better than
three alternative zero-censored Gaussian models: a standard ensemble model
output statistics approach (EMOS, \citealp{Gneiting+Raftery+Westveld:2005}), a GAMLSS
with regressors prespecified based on meteorological expertise following
\cite{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
\citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
\citep{Messner+Mayr+Zeileis:2017} as a technique for variable selection from all
82~available regressors.

@@

In weather forecasting, various factors can make it challenging to 
build a well fitting model. Complex physical processes in the 
atmosphere depend on possibly rapidly changing circumstances. 
Often it is hard to tell which of these circumstances have a 
relevant impact on the quantity of interest and what type of 
effect that can be or how they might interact. A profound 
basis of information can be obtained from numerical weather 
prediction (NWP) models as they provide predictions for a 
large number of meteorological quantities. However, while 
including too many variables in a model can lead to overfitting, 
excluding variables in advance can result in a loss of valuable 
information. Therefore, the task of selecting variables and 
specifying interactions forms the basis of building an adequate 
model. Moreover, this model has to be able to deal with a wide 
range of effects as abrupt changes can appear as well as smooth 
transitions. Especially if the considered region includes complex 
terrain, a large grid spacing of NWP models can lead to abrupt 
local effects within a single grid cell.

As explained in the introduction, distributional forests provide 
a new approach in probabilistic forecasting by embedding the ideas 
of flexible parametric modeling into the framework of tree-structured 
models. In that way, they offer features that could be particularly 
valuable in probabilistic weather forecasting in order to tackle 
the above mentioned challenges. Among these features are the ability 
to deal with a wide variety of effects and an automatic variable 
selection and detection of interactions. 
To show their strengths in practice, distributional forests are 
now applied to probabilistic precipitation forecasts in a 
mountainous region based on a large number of numerical weather 
prediction quantities. Their results are compared to those of 
other zero-censored Gaussian models including a standard ensemble 
model output approach and two GAMLSS. For one of the GAMLSS the 
regressors have been selected and interactions have been specified 
in advance based on subject-matter knowledge and experience from 
meteorology. The other one is based on a computationally more 
demanding boosting approach.

\subsection{Data}
\label{data}
The data set used for training and validation consists of observed 
precipitation sums provided by the National Hydrographical Service 
(\citealt{ehyd}) and numerical weather forecasts from the {U.S.} 
National Oceanic and Atmospheric Administration (NOAA).
Both, observations and weather forecasts are available from 1985 
trough the year 2012, focusing on July only in this article.

Observations are available on a daily basis, each representing the 
total precipitation sum fallen within 24~hours observed at 6~UTC 
at 95 stations all over Tyrol and its close surrounding. The 
observations are undergoing a rigorous quality check before made 
available to the public.  As covariates the second generation 
reforecast data set of the global ensemble forecast system (GEFS) 
is used (\citealt{Hamill+Bates+Whitaker:2013}).  This data set 
consists of an 11-member ensemble based on a fixed version of the 
numerical model and a horizontal grid-spacing of about $50~\times~50 
\text{km}^2$ initialized daily at 0~UTC from December 1984 to present 
providing forecasts on 26 hourly temporal resolution. Each of the 
11 ensemble members uses slightly different perturbed initial 
conditions trying to predict the situation specific uncertainty 
of the atmospheric state.

In particular, 14 basic forecast variables are considered with up 
to 12 variations of each of these variables being included as 
covariates. Since an ensemble of individual predictions is provided 
for each forecast variable, the possible variations include the mean, 
the standard deviation, the minimum and the maximum over all ensemble 
members, noted as ensemble mean, ensemble std. deviation, ensemble 
minimum and ensemble maximum. All of these values can be calculated 
on the sums over 24 hours (6--6 UTC of the following day) or a 6 hour 
time window. Another possible option is to consider the mean, minimum 
or maximum over 24 hours for each individual prediction and then take 
the ensemble mean or ensemble standard deviation, noted as ensemble 
mean of mean/minimal/maximal values or ensemble std. deviation of 
mean/minimal/maximal values respectively. Table~\ref{covariates} 
shows all variables together with the number of variations and 
the provided variations listed in the last column. Altogether 
82~covariates are available in this setting.

\fixme{Mention that dswrf\_mean/sprd\_min is not included?}

%%% <TH> hm, now we _are_ using a transformation model without estimating the
%%% transformation. </TH>
To remove large parts of the skewness of precipitation data, 
a power transformation has been used frequently (Box and Cox 1964). 
In literature, cubic (\citealt{Stidd:1973}) or square root 
(\citealt{Hutchinson:1998}) transformations have often been 
suggested but may vary for different climatic zones or temporal 
aggregation periods. In \cite{Stauffer+Mayr+Messner:2017} 
optimization methods have been applied to find the best transformation 
factor which resulted in a value of $\frac{1}{1.6}$. Since the 
observation data set used for this optimization contains the 
observations considered in this article the same factor is 
chosen here for the power transformation of the response variable 
\emph{tp} (total precipitation). The resulting transformed 
variable is expected to be normally distributed and left-censored 
at censoring point 0. For simplicity the same transformation has 
been chosen for the variable \emph{cape} (convective available 
potential energy).

%\newpage
\begin{table}[t!]
%\resizebox{1.4\textwidth}{!}{
\begin{minipage}{\textwidth}
\begin{tabular}{ l  c  l }
\hline
Basic covariates & {\#} & Variations\\
\hline
\emph{tp}: total precipitation            & 12  & ensemble mean of sums over 24h, \\
\hspace*{0.5cm} power transformed (by $\frac{1}{1.6}$) &    & ensemble std. deviation of sums over 24h, \\ 
\emph{cape}: convective available     &    & ensemble minimum of sums over 24h, \\
\hspace*{0.9cm} potential energy &    & ensemble maximum of sums over 24h\\
\hspace*{0.9cm} power transformed (by $\frac{1}{1.6}$)&    & \qquad all for 6--30 UTC\\
\hspace*{\fill}                            &    & ensemble mean of sums over 6h\\
\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 UTC\\
\hspace*{\fill}                            &    & ensemble std. deviation of sums over 6h\\
\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 UTC\\
\hline
\emph{dswrf}: downwards short wave      & 6 & ensemble mean of mean values, \\
\hspace*{1.13cm} radiation flux (``sunshine'') &   & ensemble mean of minimal values,\\
\emph{msl}: mean sea level pressure     &   &  ensemble mean of maximal values,\\
\emph{pwat}: preciptable water          &   & ensemble std. deviation of mean values,\\
\emph{tmax}: 2m maximum temperature     &   & ensemble std. deviation of minimal values,\\
\hspace*{1.15cm}                          &   & ensemble std. deviation of maximal values,\\
\emph{tcolc}: total column-integrated   &   & \qquad all over 6--30 UTC \\
\hspace*{0.95cm} condensate               &   & \\
\emph{t500}: temperature on 500 hPa     &   & \\
\hspace*{1cm}                             &   & \\
\emph{t700}: temperature on 700 hPa     &   & \\
\hspace*{1cm}                             &   & \\
\emph{t850}: temperature on 850 hPa     &   & \\
\hspace{1cm}                              &   & \\
\hline
\emph{tdiff500850}: temperature         & 3 & ensemble mean of difference in mean,\\
\hspace*{\fill} difference 500 to 850 hPa &   & ensemble minimum of difference in mean,\\
\emph{tdiff500700}: temperature         &   & ensemble maximum of difference in mean\\
\hspace*{\fill} difference 500 to 700 hPa &   & \qquad all over 6--30 UTC\\
\emph{tdiff700850}: temperature         &   & \\
\hspace*{\fill} difference 700 to 850 hPa &   & \\
\hline
\emph{msl{\_}diff}: mean sea level pressure & 1 & \emph{msl{\_}mean{\_}max} $-$ \emph{msl{\_}mean{\_}min}\\
\hspace{1.6cm} difference            &   & \qquad over 6--30 UTC\\
\hline
\end{tabular}
\caption[Table caption text]{Basic covariates together with the number ({\#}) and the type of variations.}
\label{covariates}
\end{minipage} 
%}
\end{table}

@@

Next to distributional forests three other methods are applied to this data set:

\begin{itemize}

\item Ensemble model output statistics 
(EMOS, \citealp{Gneiting+Raftery+Westveld:2005}): The 
location parameter and the logarithm of the scale parameter depend 
linearly on the ensemble mean forecast of total daily precipitation 
sums and the corresponding ensemble standard deviation respectively.
The choice of the logarithmic function as link function for the scale 
parameter has been made based on the results of 
\cite{Gebetsberger+Messner+Mayr:2017} where different link functions 
have been compared. Moreover, testing the identity, quadratic and 
logarithmic function as link functions in this setting supported 
this choice as the logarithmic function led to the best performance 
of the EMOS model in this application.
\item Prespecified GAMLSS: For each distribution parameter the selection 
of covariates and the specification of interactions is done in 
advance, both based on meteorological expert knowledge following 
\cite{Stauffer+Umlauf+Messner:2017}. Among the 
available 82 regressors, 9 are selected for the location parameter 
and 6 for the scale parameter. Additionally, for each distribution 
parameter one interaction term is specified in which convective 
available energy is interacting with total precipitation (with the 
ensemble mean for the location parameter and with the ensemble 
standard deviation for the scale parameter). These interactions 
might capture events with a high potential for moderate or severe 
thunderstorms with large amounts of precipitation. By putting 
extra effort in the specification of the formulas the prespecified 
GAM is provided with additional information and is therefore 
expected to profit from this advantage.
\item Boosted GAMLSS: Non-homogeneous boosting 
\citep{Messner+Mayr+Zeileis:2017} is used for an automatic 
selection of covariates \citep{Hofner+Mayr+Schmid:2016}. 
The idea of this boosting approach is to iteratively minimize 
a loss function where in each step only the coefficient leading 
to the highest improvement of the current model is updated. As 
the initial coefficients are all zero, only the most important 
covariates are included with nonzero coefficients while less 
relevant variables are ignored if the procedure is stopped before 
convergence.
\end{itemize}

%\fixme{Adding a reference to the above paragraph would be good, justifying our knowledge/experience regarding the prespecified GAMLSS. Probably Stauffer+Umlauf+Messner:2017 as I have done in the intro.}

%\fixme{A description of the boosting approach is currently missing completely.}

The selected methods and their specifications are listed in Table~\ref{model_specification}.

\begin{table}[t!]
\centering
\begin{tabular}{ l l l l }
\hline
Model & Type & Location ($\mu$) & Scale ($\log(\sigma)$) \\
\hline
\textbf{Distributional forest} & recursive    & all & all \\ 
                & partitioning &     &     \\ 
\hline
\textbf{Prespecified GAMLSS} & spline  & \emph{tp{\_}mean}, & \emph{tp{\_}sprd},\\
                            & in each & \emph{tp{\_}max},  & \emph{dswrf{\_}sprd{\_}mean},\\
 & & \emph{tp{\_}mean1218} $\ast$ & \emph{tp{\_}sprd1218} $\ast$\\ 
 & & \quad \emph{cape{\_}mean1218}, & \quad \emph{cape{\_}mean1218},\\
 & & \emph{dswrf{\_}mean{\_}mean}, & \emph{tcolc{\_}sprd{\_}mean},\\
 & & \emph{tcolc{\_}mean{\_}mean}, & \emph{tdiff500850{\_}mean}\\
 & & \emph{pwat{\_}mean{\_}mean}, & \\
 & & \emph{tdiff500850{\_}mean}, & \\
 & & \emph{msl{\_}diff} & \\
\hline
\textbf{Boosted GAMLSS} & spline  & all & all \\
                       & in each &     &     \\
\hline
\textbf{EMOS} & linear & \emph{tp{\_}mean} & \emph{tp{\_}sprd} \\
\hline
\end{tabular}
\caption[Table caption text]{Overview of models with type of 
covariate dependency and included covariates for each distribution 
parameter.}
\label{model_specification}
\end{table}

In order to compare the performance of the chosen methods, the continuous 
ranked probability score (CRPS) is calculated for each model. 
\fixme{Adding reference for CPRS}

The CRPS measures the accuracy of a probabilistic model. 
For one observation $y$ with predicted distribution 
function $F$ the CRPS value is calculated by
$$
\text{CRPS}(y, F) = \int_{-\infty}^\infty (F(z) - \mathbf{1}(y \leq z))^2 dz
$$
where \(\mathbf{1}\) is the indicator function.
%%% <TH> y > z??? </TH>

In this application the CRPS is evaluated out of sample, 
e.g. on a part of the data set that has not been considered to 
build the model. The mean CRPS over this subset of the data 
is used for validation. Since the CRPS can be seen as a 
generalization of the mean absolute error, a low value indicates 
a good fit. In applications, one big advantage in terms of 
interpretability is that the resulting value is on the scale of 
the observations.

For the purpose of comparing the models' performances with respect 
to a reference model the so-called skill score of the CRPS can be 
calculated. In that way a measure of how much a method improves the
model fit in terms of CRPS compared to the reference method is obtained.
$$
\text{SkillScore}_{\text{method}} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{reference}}}
$$

@@

\begin{figure}[t!]
\centering
\setkeys{Gin}{width=0.99\textwidth}
<<rain_axams_qqr_oosample, fig=TRUE, echo=FALSE, width = 6.5, pdf = FALSE, png = TRUE, resolution = 150>>=
# out of sample
set.seed(7)
par(mfrow = c(2,2))

qqrplot(pit_df_t, nsim = 100, main = "Distributional forest", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
qqrplot(pit_g_t, nsim = 100, main = "Prespecified GAMLSS", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
qqrplot(pit_gb_t, nsim = 100, main = "Boosted GAMLSS", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
qqrplot(pit_ml_t, nsim = 100, main = "EMOS", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
@
\caption{\label{qqr_oosample} Out-of-sample residual QQ-plots (2009--2012) 
for station Axams based on models learned on data from 1985--2008.}
\end{figure}

@@

  layout(cbind(1, 2), width = c(9, 1))
  par(mar = c(5,4,4,0.1))
  raster::image(tirol.dem, col = rev(gray.colors(100)),
                main="Stations in Tyrol", ylab = "Latitude", xlab = "Longitude", 
                xlim = c(9.8,13.2), 
                ylim = c(46.6, 47.87))
  plot(tirol.gadm, add = TRUE)
  points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)
  legend(x = 9.8, y = 47.815, pch=c(21,24,25,22), legend = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS", "EMOS"), cex = 1, bty = "n")
  text(x = 10.3, y = 47.82, labels = "Models with lowest CRPS")
  mtext("CRPS\ndifference", side=4, las = TRUE, at = c(x = 13.5, y = 47.76), line = 0.3)
  par(mar = c(0.5,0.2,0.5,2.3))
  ## legend
  plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
       xlim = c(0, 1), ylim = c(-0.2, 0.2), xaxs = "i", yaxs = "i")
  rect(0, brk[-6], 0.5, brk[-1], col = rev(clr))
  axis(4, at = brk, las = 1, mgp=c(0,-0.5,-1))

@@

par(mfrow = c(2,2))
pithist(pit_df_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Distributional forest", ylim = c(0,1.5))
pithist(pit_g_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Prespecified GAMLSS", ylim = c(0,1.5))
pithist(pit_gb_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Boosted GAMLSS", ylim = c(0,1.5))
pithist(pit_ml_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS", ylim = c(0,1.5))

>>>>

\Address{
  Lisa Schlosser, Reto Stauffer, Achim Zeileis \\
  Universit\"at Innsbruck \\
  Department of Statistics \\
  Faculty of Economics and Statistics \\
  Universit\"atsstr.~15 \\
  6020 Innsbruck, Austria \\
  E-mail: \email{Lisa.Schlosser@uibk.ac.at}, \email{Reto.Stauffer@uibk.ac.at},\\
  \email{Achim.Zeileis@R-project.org} \\
  URL: \url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/},\\
  \phantom{URL: }\url{https://retostauffer.org/},\\
  \phantom{URL: }\url{https://eeecon.uibk.ac.at/~zeileis/}\\

@@

In weather forecasting, these properties are appealing in mountainous regions
and complex terrain where small-scale effects are not well-resolved in the NWP
outputs. As these are computed at a coarser scale, there may be abrupt local
effects within a single NWP grid cell. To illustrate this in practice,
precipitation forecasts are obtained with distributional forests at 95
meteorological stations in a mountainous region in the Alps, covering mainly
Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{map}).
More specifically, a zero-censored Gaussian
distribution is employed to model 24-hour total precipitation so that the
zero-censored point mass describes the probability of no precipitation on a
given day (see Figure~\ref{axams_onestation}). Forecasts for July are
established based on data from the same month in the years 1985--2012 along with
82~covariates derived from a wide range of NWP ensemble outputs. As
Figure~\ref{axams_onestation} shows, the station-wise models yield a full
distributional forecast for each day -- here for the same day (July~24) at one
station (Axams) over four years (2009--2012) -- based on the previous 24~years
as training data. The corresponding observations conform reasonably well with
the predictions. In Section~\ref{ProbabilisticForecasting} we investigate the
performance of distributional forests in this forecasting task in more detail.
It is shown that they perform at least on par and sometimes clearly better than
three alternative zero-censored Gaussian models: a standard ensemble model
output statistics approach (EMOS, \citealp{Gneiting+Raftery+Westveld:2005}), a GAMLSS
with regressors prespecified based on meteorological expertise following
\cite{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
\citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
\citep{Messner+Mayr+Zeileis:2017} as a technique for variable selection from all
82~available regressors.

@@

Many weather forecasting models leverage the strengths of modern
numerical weather prediction (NWP) systems \citep[see][]{Bauer+Thorpe+Brunet:2015}
and produce high-quality forecasts adapted to specific sites/locations
through distributional regression models like EMOS
(ensemble model output statistics, \citealp{Gneiting+Raftery+Westveld:2005}).
In the case of precipiation forecasting, EMOS explains past precipitation
sums by the NWP output ``total precipitation'' (\emph{tp}), using its
ensemble mean as the predictor variable for the location parameter and
the ensemble standard deviation for the scale parameter.

While this approach alone is already highly effective in the plains,
it typically does not perform as well in complex terrain due to unresolved
effects in the NWP system. For example, in the Tyrolean Alps -- considered
in the following application case study -- the NWP grid cells of $50 \times 50$
km$^2$ are too coarse to capture mountains, valleys, etc. Therefore,
it is often possible to substantially improve the EMOS by including
further predictor variables, either from local observations or from
other NWP outputs. Unfortunately, it is typically unclear which
variables are relevant for improving the predictions. Simply including
all available variables may be computationally burdensome and
can lead to overfitting but, on the other hand, excluding too many variables
may result in a loss of valuable information. Therefore, selecting
the relevant variables and interactions is crucial for improving the
forecasting model.

In the following, it is illustrated how distributional regression forests
can solve this problem of automatically selecting relevant variables,
interactions, and potentially nonlinear effects. For fitting the forest
only the response distribution and the list of potential predictor variables
need to be specified (along with a few algorithmic details) and then
the model fit is determined by the forest in a data-driven way. Here, we
employ a zero-censored Gaussian distribution and 82~predictor variables
computed from ensemble means and spreads of various NWP outputs.
The predictive performance of the forest is compared to three other
zero-censored Gaussian models: (a) a standard basic EMOS, (b) a GAMLSS
with prespecified effects and interactions based on meteorological
knowedge/experience, and (c) a boosted GAMLSS with automatic
selection of smooth additive terms based on all 82~predictor variables.

\subsection{Data}
\label{data}

\begin{table}[t!]
%\resizebox{1.4\textwidth}{!}{
\begin{minipage}{\textwidth}
\begin{tabular}{ l  c  l }
\hline
Basic covariates & {\#} & Variations\\
\hline
\emph{tp}: total precipitation            & 12  & ensemble mean of sums over 24h, \\
\hspace*{0.5cm} power transformed (by $\frac{1}{1.6}$) &    & ensemble std. deviation of sums over 24h, \\ 
\emph{cape}: convective available     &    & ensemble minimum of sums over 24h, \\
\hspace*{0.9cm} potential energy &    & ensemble maximum of sums over 24h\\
\hspace*{0.9cm} power transformed (by $\frac{1}{1.6}$)&    & \qquad all for 6--30 UTC\\
\hspace*{\fill}                            &    & ensemble mean of sums over 6h\\
\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 UTC\\
\hspace*{\fill}                            &    & ensemble std. deviation of sums over 6h\\
\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 UTC\\
\hline
\emph{dswrf}: downwards short wave      & 6 & ensemble mean of mean values, \\
\hspace*{1.13cm} radiation flux (``sunshine'') &   & ensemble mean of minimal values,\\
\emph{msl}: mean sea level pressure     &   &  ensemble mean of maximal values,\\
\emph{pwat}: preciptable water          &   & ensemble std. deviation of mean values,\\
\emph{tmax}: 2m maximum temperature     &   & ensemble std. deviation of minimal values,\\
\hspace*{1.15cm}                          &   & ensemble std. deviation of maximal values,\\
\emph{tcolc}: total column-integrated   &   & \qquad all over 6--30 UTC \\
\hspace*{0.95cm} condensate               &   & \\
\emph{t500}: temperature on 500 hPa     &   & \\
\hspace*{1cm}                             &   & \\
\emph{t700}: temperature on 700 hPa     &   & \\
\hspace*{1cm}                             &   & \\
\emph{t850}: temperature on 850 hPa     &   & \\
\hspace{1cm}                              &   & \\
\hline
\emph{tdiff500850}: temperature         & 3 & ensemble mean of difference in mean,\\
\hspace*{\fill} difference 500 to 850 hPa &   & ensemble minimum of difference in mean,\\
\emph{tdiff500700}: temperature         &   & ensemble maximum of difference in mean\\
\hspace*{\fill} difference 500 to 700 hPa &   & \qquad all over 6--30 UTC\\
\emph{tdiff700850}: temperature         &   & \\
\hspace*{\fill} difference 700 to 850 hPa &   & \\
\hline
\emph{msl{\_}diff}: mean sea level pressure & 1 & \emph{msl{\_}mean{\_}max} $-$ \emph{msl{\_}mean{\_}min}\\
\hspace{1.6cm} difference            &   & \qquad over 6--30 UTC\\
\hline
\end{tabular}
\caption[Table caption text]{Basic covariates together with the number ({\#}) and the type of variations.}
\label{covariates}
\end{minipage} 
%}
\end{table}

Training and validation data consist of observed 
daily precipitation sums provided by the National Hydrographical Service 
(\citealt{ehyd}) and numerical weather forecasts from the U.S.~National
Oceanic and Atmospheric Administration (NOAA).
Both, observations and forecasts are available for 1985--2012 and
the analysis is exemplified using July as the month with the most precipitation
in Tyrol.

Observations are obtained for 95~stations all over Tyrol and surroundings,
providing 24~hour precipitation sums measured at 6~UTC and rigorously
quality-checked by the National Hydrographical Service. NWP outputs
are obtained from the second generation reforecast data set of the
global ensemble forecast system \citep[GEFS,][]{Hamill+Bates+Whitaker:2013}. 
This data set consists of an 11-member ensemble based on a fixed version of the 
numerical model and a horizontal grid spacing of about $50 \times 50$ 
km$^2$ initialized daily at 0~UTC from December 1984 to present 
providing forecasts on 26 hourly temporal resolution. Each of the 
11 ensemble members uses slightly different perturbed initial 
conditions trying to predict the situation specific uncertainty 
of the atmospheric state.

From the GEFS outputs 14 basic forecast variables are considered with up 
to 12 variations of each of these variables such as mean/maximum/minimum
or different aggregation time ranges, etc. A detailed overview is
provided in Table~\ref{covariates}, yielding 82~predictor variables in total.
\fixme{Mention that dswrf\_mean/sprd\_min is not included?}

To remove large parts of the skewness of precipitation data, 
a power transformation \citep{Box+Cox:1964} is often applied, e.g., using
cubic \citep{Stidd:1973} or square root \citep{Hutchinson:1998} transformations.
However, the power parameter may vary for different climatic zones or temporal 
aggregation periods and hence we follow \cite{Stauffer+Mayr+Messner:2017} 
in their choice of $\frac{1}{1.6}$ as a suitable power parameter for the
region of Tyrol. The same power transformation is applied to both the
observed precipitation sums and the NWP outputs ``total precipitation'' (\emph{tp}) and
``convective available potential energy'' (\emph{cape}).

@@

\begin{table}[t!]
\centering
\begin{tabular}{ l l l l }
\hline
Model & Type & Location ($\mu$) & Scale ($\log(\sigma)$)                                  \\ \hline
Distributional forest & recursive    & all                & all                           \\ 
                      & partitioning &                    &                               \\ \hline
EMOS                  & linear       & \emph{tp{\_}mean}  & \emph{tp{\_}sprd}             \\ \hline
Prespecified GAMLSS   & spline       & \emph{tp{\_}mean}, & \emph{tp{\_}sprd},            \\
                      & in each      & \emph{tp{\_}max},  & \emph{dswrf{\_}sprd{\_}mean}, \\
 & & \emph{tp{\_}mean1218} $\ast$ & \emph{tp{\_}sprd1218} $\ast$\\ 
 & & \quad \emph{cape{\_}mean1218}, & \quad \emph{cape{\_}mean1218},\\
 & & \emph{dswrf{\_}mean{\_}mean}, & \emph{tcolc{\_}sprd{\_}mean},\\
 & & \emph{tcolc{\_}mean{\_}mean}, & \emph{tdiff500850{\_}mean}\\
 & & \emph{pwat{\_}mean{\_}mean}, & \\
 & & \emph{tdiff500850{\_}mean}, & \\
 & & \emph{msl{\_}diff} & \\ \hline
Boosted GAMLSS        & spline  & all & all \\
                      & in each &     &     \\ \hline
\end{tabular}
\caption[Table caption text]{Overview of models with type of 
covariate dependency and included covariates for each distribution 
parameter.}
\label{model_specification}
\end{table}

The following zero-censored Gaussian regression models are employed
in the empirical case study, see also Table~\ref{model_specification} for
further details:
%
\begin{itemize}

\item \emph{Distributional forest:} All 82~predictor variables are
  considered for learning a forest of 100 trees. Bootstrap
  sampling is employed for each tree using a third of the predictors
  in each split of the tree (``mtry''). Parameters are estimated by
  adaptive local likelihood based on the forest weights, as described
  in Section~\ref{Methodology}.

\item \emph{EMOS:} Standard ensemble model output statistics are used
  with the total precipitation ensemble mean as regressor in the location
  submodel and the ensemble standard deviation in the scale submodel.
  The parameters are estimated by maximum likelihood, using an identity
  link for the location and a log link for the scale
  \citep[following the advice of][]{Gebetsberger+Messner+Mayr:2017}.

\item \emph{Prespecified GAMLSS:} Smooth additive splines are selected for
  the most relevant predictors, based on meteorological expert knowledge
  following \cite{Stauffer+Umlauf+Messner:2017}. More specifically, based
  on the 82~available variables, 8 terms are included in the location
  submodel and 6 in the scale submodel. Both involve an interaction of
  \emph{tp} and \emph{cape} in the afternoon (between 12 and 18 UTC)
  to capture the potential for thunderstorms that frequently occur in
  summer afternoons in the Alps. The model is estimated by maximum
  penalized likelihood using a backfitting algorithm \citep{Stasinopoulos+Rigby:2007}.

\item \emph{Boosted GAMLSS:} Smooth additive splines are selected
  automatically from all 82~available variables, using non-cyclic boosting
  for parameter estimation \citep{Hofner+Mayr+Schmid:2016,Messner+Mayr+Zeileis:2017}.
  This updates the predictor terms for the location or scale submodels iteratively
  by maximizing the log-likelihood only for the variable yielding the
  highest improvement. The iteration stops early -- before fully maximizing
  the in-sample likelihood -- based on a (computationally intensive)
  out-of-bag bootstrap estimate of the log-likelihood. The grid considered for
  the number of boosting iterations (``mstop'') is: $50, 75, \dots, 975, 1000$.

The predictive performance in terms of full probabilistic forecasts is
assessed using the continuous ranked probability score (CRPS, \fixme{reference}).
For each of the models this assesses the discrepancy of the
predicted distribution function $F$ from the observation $y$ by
$$
\text{CRPS}(y, F) = \int_{-\infty}^\infty (F(z) - \mathbf{1}(y \leq z))^2 dz
$$
where \(\mathbf{1}\) is the indicator function. In the subsequent
applications, the mean CRPS is always evaluated out of sample, 
either using cross-validation or a hold-out data set (2009--2012)
that was not used for learning (1985--2008). CRPS is a proper
scoring rule \fixme{citation} and lower values indicate a better fit.

To assess differences in the improvement of the forests and GAMLSS
models over the basic EMOS, a CRPS-based skill score with EMOS as
the reference method is computed as well:
$$
\text{SkillScore}_{\text{method}} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{EMOS}}}.
$$

@@

\begin{figure}[t!]
\centering
\setkeys{Gin}{width=0.99\textwidth}
<<rain_axams_qqr_oosample, fig=TRUE, echo=FALSE, width = 6.5, pdf = FALSE, png = TRUE, resolution = 150>>=
# out of sample
set.seed(7)
par(mfrow = c(2,2))
qqrplot(pit_df_t, nsim = 100, main = "Distributional forest", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
qqrplot(pit_ml_t, nsim = 100, main = "EMOS", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
qqrplot(pit_g_t, nsim = 100, main = "Prespecified GAMLSS", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
qqrplot(pit_gb_t, nsim = 100, main = "Boosted GAMLSS", ylim = c(-5,5), col = gray(0.04, alpha = 0.01), pch = 19)
@
\caption{\label{qqr_oosample} Out-of-sample residual QQ-plots (2009--2012) 
for station Axams based on models learned on data from 1985--2008.}
\end{figure}

@@

  layout(cbind(1, 2), width = c(9, 1))
  par(mar = c(5,4,4,0.1))
  raster::image(tirol.dem, col = rev(gray.colors(100)),
                main="Stations in Tyrol", ylab = "Latitude", xlab = "Longitude", 
                xlim = c(9.8,13.2), 
                ylim = c(46.6, 47.87))
  plot(tirol.gadm, add = TRUE)
  points(sp, pch = c(21, 24, 25, 22)[bst], bg = clr[grp], col = "black", las = 1, cex = 1.5)
  legend(x = 9.8, y = 47.815, pch = c(21, 24, 25, 22), legend = c("Distributional forest", "Prespecified GAMLSS", "Boosted GAMLSS", "EMOS"), cex = 1, bty = "n")
  text(x = 10.3, y = 47.82, labels = "Models with lowest CRPS")
  mtext("CRPS\ndifference", side=4, las = TRUE, at = c(x = 13.5, y = 47.76), line = 0.3)
  par(mar = c(0.5,0.2,0.5,2.3))
  ## legend
  plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "",
       xlim = c(0, 1), ylim = c(-0.2, 0.2), xaxs = "i", yaxs = "i")
  rect(0, brk[-6], 0.5, brk[-1], col = rev(clr))
  axis(4, at = brk, las = 1, mgp=c(0,-0.5,-1))

@@

par(mfrow = c(2,2))
pithist(pit_df_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Distributional forest", ylim = c(0,1.5))
pithist(pit_ml_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "EMOS", ylim = c(0,1.5))
pithist(pit_g_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Prespecified GAMLSS", ylim = c(0,1.5))
pithist(pit_gb_t, nsim = 1000, breaks = seq(0, 1, length.out = 9), main = "Boosted GAMLSS", ylim = c(0,1.5))
